---
title: "p8105_HW2_cw3555"
author: "Eunice Wang"
date: "2023-10-04"
output: github_document
---

```{r,message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
First, clean the data in _pols-month.csv_.

```{r}
pols_df =
  read_csv("./fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names() |> 
  separate(mon,sep = "-", into = c("year","month","day")) |> 
  mutate(
    month = case_match(
      month,
      "01" ~ "Jan",
      "02" ~ "Feb",
      "03" ~ "Mar",
      "04" ~ "Apr",
      "05" ~ "May",
      "06" ~ "Jun",
      "07" ~ "Jul",
      "08" ~ "Aug",
      "09" ~ "Sep",
      "10" ~ "Oct",
      "11" ~ "Nov",
      "12" ~ "Dec"
    ) ) |> 
  mutate(
    president = case_match(
      prez_gop,
      1 ~ "gop",
      2 ~ "gop",
      0 ~ "dem"
  )) |> 
   select(-prez_dem, -prez_gop,-day)
```

* Cleaned pol-month dataset contains `r nrow(pols_df)` observations of `r ncol(pols_df)` variables related to the number of national politicians who are democratic or republican at any given time from `r min(pols_df$year)` to `r max(pols_df$year)`.The variable _president_ represents their partisanship.

Second, clean the data in _snp.csv_ using a similar process to the above.

```{r}
snp_df =
  read_csv("./fivethirtyeight_datasets/snp.csv") |>
  janitor::clean_names() |> 
  separate(date,sep = "/", into = c("month","day","year")) |> 
  mutate(
    year = as.numeric(year),
    year = ifelse(year>=50, 1900+year, 2000+year)
  ) |> 
  relocate(year,month) |> 
  mutate(
    month = case_match(
      month,
      "1" ~ "Jan",
      "2" ~ "Feb",
      "3" ~ "Mar",
      "4" ~ "Apr",
      "5" ~ "May",
      "6" ~ "Jun",
      "7" ~ "Jul",
      "8" ~ "Aug",
      "9" ~ "Sep",
      "10" ~ "Oct",
      "11" ~ "Nov",
      "12" ~ "Dec"
    ) ) |> 
  select(-day)
```

* Cleaned snp dataset contains `r nrow(snp_df)` observations of `r ncol(snp_df)` variables related to Standard & Poor’s stock market index (S&P), often used as a representative measure of stock market as a whole,from `r min(snp_df$year)` to `r max(snp_df$year)`. The varibale _close_ represents the closing values of the S&P stock index on the associated date.

Third, tidy the _unemployment data_ so that it can be merged with the previous datasets.

```{r}
unemploy_df =
  read_csv("./fivethirtyeight_datasets/unemployment.csv") |>
    pivot_longer(
    Jan:Dec,
    names_to = "month", 
    values_to = "unemploy_rate"
  ) |> 
    janitor::clean_names() |> 
    mutate(year = as.numeric(year)) 
```

* Cleaned unemployment dataset contains `r nrow(unemploy_df)` observations of `r ncol(unemploy_df)` variables related to the unemployment rate in any given month from `r min(unemploy_df$year)` to `r max(unemploy_df$year)`.

Finally,join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
result_df =
  pols_df |> 
  mutate(year = as.numeric(year)) |> 
  left_join(snp_df, by = join_by(year,month)) |> 
  left_join(unemploy_df, by = join_by(year,month))
```

* The result dataset contains `r nrow(result_df)` observations of `r ncol(result_df)` variables related all three parts of information above (including partisianship, S&P, unemployment rate) in any given month from `r min(result_df$year)` to `r max(result_df$year)`.

## Problem 2

First, read and clean the Mr. Trash Wheel sheet
```{r}
mr_df =
  readxl::read_excel("trashwheel_datasets/202309 Trash Wheel Collection Data.xlsx",sheet = "Mr. Trash Wheel",range = "A2:N587") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    homes_powered = weight_tons * 500 /30,
    name = "mr"
  )
```

Next, use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda

```{r}
professor_df =
  readxl::read_excel("trashwheel_datasets/202309 Trash Wheel Collection Data.xlsx",sheet = "Professor Trash Wheel",range = "A2:M109") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    homes_powered = weight_tons * 500 /30,
    name = "professor"
  )
```

```{r}
gwynnda_df =
  readxl::read_excel("trashwheel_datasets/202309 Trash Wheel Collection Data.xlsx",sheet = "Gwynnda Trash Wheel",range = "A2:L159") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    homes_powered = weight_tons * 500 /30,
    name = "gwynnda"
  )
```

Finally, combine these three dataset to produce a single tidy dataset

```{r}
trashwheel_df =
  bind_rows(mr_df,professor_df,gwynnda_df) 
```

* The resulting `trashwheel_df` dataset is a combination of 3 datasets _Mr. Trash Wheel_, _Professor Trash Wheel_, and  _Gwynnda Trash Wheel_ . In total, the `trashwheel_df` dataset  contains`r nrow(trashwheel_df)` observations of `r ncol(trashwheel_df)` variables, among which the `name` variable represent the trash wheel's name. The resulting dataset contains `r nrow(filter(trashwheel_df, name == "mr"))` observations from _Mr. Trash Wheel_ dataset, `r nrow(filter(trashwheel_df, name == "professor"))` observations from _Professor Trash Wheel_ dataset, and `r nrow(filter(trashwheel_df, name == "gwynnda"))` observations from _Gwynnda Trash Wheel_ dataset.
The resulting dataset contains key variables such as data (following `month`, `year`, `date` variables), and the workload of each trashwheel at a specific data (including the weight and volume of trash it collected `weight_tons` and `volume_cubic_yards`). It also contains amount of specific kind of trash that each trashwheel collected: `plastic_bottles`,`polystyrene`,`cigarette_butts`, `glass_bottles`(which is not collected by Gwynnda Trash Wheel), `plastic_bags`, `wrappers`, `sports_balls` (which is only collected by Mr.Trash Wheel). It also contains the (approximate) number of homes powered, `homes_powered`.

* Q: For available data, what was the total weight of trash collected by Professor Trash Wheel? 
* A: The total weight of trash collected by Professor Trash Wheel is 
`r sum(pull(professor_df,weight_tons))`
* Q:What was the total number of cigarette butts collected by Gwynnda in July of 2021?
* A: The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum(pull(filter(gwynnda_df, year == 2021, month == "July"),cigarette_butts))`

## Problem 3

Import, clean, and tidy the dataset of baseline demographics. Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline).

```{r}
baseline_df =
  read_csv("./data_mci/MCI_baseline.csv",skip=1) |>
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      1 ~ "Male",
      0 ~ "Female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "APOE4 carrier",
      0 ~ "APOE4 non-carrier"
    ) ) |> 
    filter(current_age < age_at_onset| age_at_onset == "." )
```

* To clean and tidy the dataset of baseline demographics, I first skipped the first row containing explanation of each variable of the dataset when I imported and then clean the names for convenience afterwards. Next, I use `case_match` function to change numeric values of variables `sex` and `apoe4` to characters of its meaning which is more appropriate to read. Finally, I use `filter` function to remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline), which is either `current_age` less than `age_at_onset` or with "." in `age_at_onset`. 
* After cleaning, the result dataset contains `r nrow(baseline_df)` observations of `r ncol(baseline_df)` variables. 

* Q: How many participants were recruited, and of these how many develop MCI? 
* A: `r nrow(read_csv("./data_mci/MCI_baseline.csv",skip=1))` participants were recruited, and `r nrow(filter(baseline_df, age_at_onset >0))` participants develop MCI.
* Q: What is the average baseline age? 
* A: The average baseline age is `r mean(pull(baseline_df, current_age))`.
* Q: What proportion of women in the study are APOE4 carriers?
* A: `r nrow(filter(baseline_df, sex=="Female", apoe4=="APOE4 carrier")) / nrow (filter(baseline_df, sex=="Female") ) * 100`% of women in the study are APOE4 carriers.

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

```{r}
amyloid_df =
  read_csv("./data_mci/mci_amyloid.csv",skip=1) |>
  janitor::clean_names() |> 
  pivot_longer(
    baseline: time_8,
    names_to = "time_elapsed",
    values_to = "amyloid_ratio",
    names_prefix = "time_"
  ) |> 
 mutate(
   time_elapsed = replace(time_elapsed, time_elapsed == "baseline",0),
   amyloid_ratio = replace(amyloid_ratio, amyloid_ratio == "Na", NA) 
   )|> 
  rename(id = study_id)
```

* To clean and tidy the dataset of f longitudinally observed biomarker values, I first skipped the first row containing explanation of each variable of the dataset when I imported and then clean the names for convenience afterwards. Next, I changed the 5 variables(`baseline`, `time2`,`time4` ,`time6`,`time8`) measuring the amyloid 42/40 ratio at difference times since the study baseline to two variables measuring `time_elapsed` and `amyloid_ratio`. For convenience, I change "baseline" to 0 to align with other representation of specific time. Because the original data contains an observation with `Na` in `baseline`, I change it to `NA` for later convenience. Finally, I change the column name `study id` to `id` for convenient match of the other dataset afterwards.
* After cleaning, the result dataset contains `r nrow(amyloid_df)` observations of `r ncol(amyloid_df)` variables. 

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings.

```{r}
baseline_only_df =
  anti_join(baseline_df, amyloid_df, by="id")
id_baseline_only =
  select(baseline_only_df, id)
amyloid_only_df =
  anti_join(amyloid_df, baseline_df, by="id")
id_amyloid_only_multiple =
  select(amyloid_only_df, id)
id_amyloid_only =
  unique(id_amyloid_only_multiple, incomparables = FALSE)
```


* There are `r nrow(baseline_only_df)` participants only appear in the baseline dataset (these participants are `id_baseline_only` with id 4,49,92,179,268,304,389 and 412),  while there are `r nrow(amyloid_only_df)/5` participants(one id appears five times) participants only appear in the amyloid datase (these participants are `id_amyloid_only` with id 72,234,283,380,484,485,486,487,488,489,490,491,492,493,494 and 495 

Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.

```{r}
both_joined_df =
  inner_join(baseline_df, amyloid_df, join_by(id))
write_csv(both_joined_df, "data_mci/AD_biomarker.csv")
```

* The resulting dataset `both_joined_df` contains the data of `r nrow(both_joined_df)` observations, `r nrow(both_joined_df)/5` participants(because we tidy the `amyloid_df`to make each participant with 5 observations related with `time_elapsed`,each id appears 5 times) that appear in both datasets, and the observation of `r ncol(both_joined_df)` variables. These variables including the basic demographic information of the participants (their`sex`,`current_age`,`education`). Their age of MCI onset `age_at_onset` and whether they are apoe4 carrier `apoe4`. It also includes the time (in years) elapsed since the study baseline to the visit where biomarker amyloid 42/40 ratio was measured: `baseline`, `time2`,`time4` ,`time6`,`time8`, and the ratio measured `amyloid_ratio`. 
* There are `r nrow(filter(both_joined_df, sex == "Female"))/5` participants who are female, and `r nrow(filter(both_joined_df, sex == "Male"))/5` participants who are male.
* There are `r nrow(filter(both_joined_df, apoe4 == "APOE4 carrier"))/5` participants who are APOE4 carrier, and `r nrow(filter(both_joined_df, apoe4 == "APOE4 non-carrier"))/5` participants who are APOE4 non-carrier.
* There are `r nrow(filter(both_joined_df, age_at_onset >0))/5` participants who develop MCI
* The resulting dataset is exported as `AD_biomarker.csv` in "data_mci".
