---
title: "p8105_HW2_cw3555"
author: "Eunice Wang"
date: "2023-10-04"
output: github_document
---

```{r,message=FALSE}
library(tidyverse)
```

## Problem 1
First, clean the data in _pols-month.csv_.

```{r}
pols_df =
  read_csv("./fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names() |> 
  separate(mon,sep = "-", into = c("year","month","day")) |> 
  mutate(
    month = case_match(
      month,
      "01" ~ "Jan",
      "02" ~ "Feb",
      "03" ~ "Mar",
      "04" ~ "Apr",
      "05" ~ "May",
      "06" ~ "Jun",
      "07" ~ "Jul",
      "08" ~ "Aug",
      "09" ~ "Sep",
      "10" ~ "Oct",
      "11" ~ "Nov",
      "12" ~ "Dec"
    ) ) |> 
  mutate(
    president = case_match(
      prez_gop,
      1 ~ "gop",
      2 ~ "gop",
      0 ~ "dem"
  )) |> 
   select(-prez_dem, -prez_gop,-day)
```

Cleaned pol-month data set contains `r nrow(pols_df)` observations of `r ncol(pols_df)` variables related to the number of national politicians who are democratic or republican at any given time from `r min(pols_df$year)` to `r max(pols_df$year)`.The variable _president_ represents their partisanship.

Second, clean the data in _snp.csv_ using a similar process to the above.

```{r}
snp_df =
  read_csv("./fivethirtyeight_datasets/snp.csv") |>
  janitor::clean_names() |> 
  separate(date,sep = "/", into = c("month","day","year")) |> 
  mutate(
    year = as.numeric(year),
    year = ifelse(year>50, 1900+year, 2000+year)
  ) |> 
  relocate(year,month) |> 
  mutate(
    month = case_match(
      month,
      "1" ~ "Jan",
      "2" ~ "Feb",
      "3" ~ "Mar",
      "4" ~ "Apr",
      "5" ~ "May",
      "6" ~ "Jun",
      "7" ~ "Jul",
      "8" ~ "Aug",
      "9" ~ "Sep",
      "10" ~ "Oct",
      "11" ~ "Nov",
      "12" ~ "Dec"
    ) ) |> 
  select(-day)
```

Cleaned snp data set contains `r nrow(snp_df)` observations of `r ncol(snp_df)` variables related to Standard & Poor’s stock market index (S&P), often used as a representative measure of stock market as a whole,from `r min(snp_df$year)` to `r max(snp_df$year)`. The varibale _close_ represents the closing values of the S&P stock index on the associated date.

Third, tidy the _unemployment data_ so that it can be merged with the previous datasets.

```{r}
unemploy_df =
  read_csv("./fivethirtyeight_datasets/unemployment.csv") |>
    pivot_longer(
    Jan:Dec,
    names_to = "month", 
    values_to = "unemploy_rate"
  ) |> 
    janitor::clean_names() |> 
    mutate(year = as.numeric(year)) 
```

Cleaned unemployment data set contains `r nrow(unemploy_df)` observations of `r ncol(unemploy_df)` variables related to the unemployment rate in any given month from `r min(unemploy_df$year)` to `r max(unemploy_df$year)`.

Finally,join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
result_df =
  pols_df |> 
  mutate(year = as.numeric(year)) |> 
  left_join(snp_df, by = join_by(year,month)) |> 
  left_join(unemploy_df, by = join_by(year,month))
```

The result data set contains `r nrow(result_df)` observations of `r ncol(result_df)` variables related all three parts of information above (including partisianship, S&P, unemployment rate) in any given month from `r min(result_df$year)` to `r max(result_df$year)`.

